A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-05-04 paper.tex
Align figure 2 properly
Minor edits

The reactive synthesis problem is EXPTIME-complete so
naive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

We evaluate our method on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).  While our solver solves fewer total instances than competitors,
it solves the largest number of unique instances, i.e., instances that could not be solved by 
any other solver.  These results confirm that there exist classes of problems that are
hard for traditional synthesis techniques, but can be efficiently solved by our method.
While further performance improvements are clearly needed, in its current state the method 
can be used in a portfolio solver to increase the number of tractable instances.

@@

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. The set of winning states has no compact representation, which
makes the problem hard for BDD solvers. In Section~3 we will outline how this
example can be solved without enumerating all winning states.

@@

@@

        \begin{tikzpicture}[level distance = 5mm]

@@

        \captionsetup{width=.8\textwidth}
        \caption{Environment-winning trace}

@@

        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]

@@

        \captionsetup{width=.8\textwidth}

@@

        \captionsetup{width=.8\textwidth}
        \caption{Controller-winning trace}

@@

    \end{subfigure}%

@@

    To solve our example, the algorithm begins by invoking a SAT solver to find
    a trace of length $k$ (here we consider $k=3$) from the initial state to an
    error state (see Figure~\ref{fig:trace}).  Given a trace we make an initial
    conjecture that the first action, \texttt{req=1}, is winning for the
    environment. This conjecture is captured in the abstract game tree show in
    Figure~\ref{fig:agt}.  Next the solver attempts to find a counterexample to
    this conjecture by finding a trace that avoids the error states with the
    first environment action fixed to \texttt{req=1}, i.e. by playing the
    abstract game on behalf of the controller.

@@

    One option is using a BDD solver, working backwards from the error set, to
    find all losing states. One iteration of this operation on our example
    would give the set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2
    \land ( \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$.  In the general case there is no compact
    representation of the losing set, so we try to avoid computing it by
    employing interpolation instead. The potential benefit of interpolation is
    that an interpolant can approximate the losing states in such a way that
    there is just enough information to prove realisability. It is also cheap
    to compute an interpolant from the proof of unsatisfiability of a SAT
    solver.

@@

We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose
children are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree
at $n$ such that both slices $T_1$ and $T_2$ contain a copy of $n$ (line~4).
Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$ except $n$'s
children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its
children.  There is no candidate strategy for $T$ so $s \land
\textsc{\textoverline{treeFormula}}(k, T)$ is unsatisfiable.  By construction,
$\textsc{\textoverline{treeFormula}}(k, T) \equiv
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ and hence $s \land
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ is also unsatisfiable. This enables
the construction of an \emph{interpolant} that captures losing states at node
$n$.

@@

@@

We have discovered a set $\mathcal{I}$ of states losing for the environment.
Environment-losing states are only losing for a particular bound: given that
there does not exist an environment strategy that forces the game into an error
state in $k$ rounds or less; there may still exist a longer environment-winning
strategy.  We therefore record learned environment-losing states along with
associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing (\textit{bad}) for the controller, indexed by
bound $k$.  $B^m[k]$ are initialised to $\top$ for all $k$.  Whenever an
environment-losing set $\mathcal{I}$ is discovered for a node $n$ with bound
$\textsc{height}(k, n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is
subtracted from $B^m[i]$, for all $i$ less than or equal to the bound
(lines~14--16).

@@

We also present a cactus plot of the number of instances solved over time (Figure~\ref{fig:cactus}). For
reference we plotted our results along with \textsc{Demiurge}~\cite{bloem2014},
the only SAT-based tool in the competition, and the winner of the sequential
realisability track \textsc{Simple BDD Solver 2}~\cite{walker2014}. We were
able to run \textsc{Demiurge} on our hardware but show competition results for
\textsc{Simple BDD Solver 2}.

>>>>

The reactive synthesis problem is EXPTIME-complete so
na\"ive algorithms are infeasible on anything but simple systems.
% In this way the environment is equivalent to an existential search of
%the game and the controller is universal. Much of the complexity of reactive
%synthesis stems from managing the interactions of the alternating quantifiers
%on the state space of the game.
There are several techniques that aim to mitigate this complexity by
representing states and transitions of the system symbolically~\cite{Piterman_PS_06,bloem2014,morgenstern2013}. 
These techniques incrementally construct a symbolic representation of the set of 
discovered winning states of the game.  The downside of this approach is that 
keeping track of all discovered winning states can lead to a space explosion even
when using efficient symbolic representation such as BDDs or CNFs.

@@

We evaluate our method on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).  While our solver solves fewer total instances than competitors,
it solves the largest number of unique instances, i.e., instances that could not be solved by 
any other solver.  These results confirm that there exist classes of problems that are
hard for traditional synthesis techniques, but can be efficiently solved by our
method.  While further performance improvements are clearly needs, in its
current state the method may be a worthwhile addition to a portfolio solver.
%%%While further performance improvements are clearly needed, in its current state the method 
%%%can be used in a portfolio solver to increase the number of tractable instances.

@@

This example is the $n=2$ instance of the more general problem of an arbiter of
$n$ resources. The set of winning states has no compact representation, which
makes the problem hard for BDD solvers. In Section~3 we will outline how the
unbounded game can be solved without enumerating all winning states.

@@

    \captionsetup[subfigure]{width=\textwidth,justification=raggedleft}

@@

        \begin{minipage}[t][3.8cm][t]{\textwidth}
        \begin{tikzpicture}[level distance = 5mm,baseline]

@@

    \end{minipage}
        \caption{Environment winning trace}

@@

        \begin{minipage}[t][3.8cm][t]{\textwidth}
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm,baseline]

@@

        \end{minipage}

@@

        \end{minipage}
        \caption{Controller winning trace}

@@

@@

    To solve our example, the algorithm begins by invoking a SAT solver to find
    a trace of length $k$ (here we consider $k=3$) from the initial state to an
    error state (see Figure~\ref{fig:trace}).  Given a trace we make an initial
    conjecture that the first action, \texttt{req=1}, is winning for the
    environment. This conjecture is captured in the abstract game tree shown in
    Figure~\ref{fig:agt}.  Next the solver attempts to find a counterexample to
    this conjecture by finding a trace that avoids the error states with the
    first environment action fixed to \texttt{req=1}, i.e. by playing the
    abstract game on behalf of the controller.

@@

    One option is using a BDD solver, working backwards from the error set, to
    find all losing states. One iteration of this operation on our example
    would give the set: $\texttt{nrequests} = 3 \lor (\texttt{nrequests} = 2
    \land ( \texttt{resource0} = 0 \lor \texttt{resource1} = 0)) \lor
    (\texttt{nrequests} = 1 \land (\texttt{resource0} = 0 \land
    \texttt{resource1} = 0))$.  In the general case there is no compact
    representation of the losing set, so we try to avoid computing it by
    employing interpolation instead. The potential benefit of interpolation is
    that an interpolant can approximate the losing states in such a way that
    there is just enough information to prove realisability. It is also cheap
    to compute an interpolant using a SAT solver.

@@

@@

We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose
children are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree
at $n$ such that both slices $T_1$ and $T_2$ contain a copy of $n$ (line~4).
Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$ except $n$'s
children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its
children.  There is no candidate strategy for $T$ so $s \land
\textsc{\textoverline{treeFormula}}(k, T)$ is unsatisfiable.  By construction,
$\textsc{\textoverline{treeFormula}}(k, T) \equiv
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ and hence $s \land
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ is also unsatisfiable.

%%%This enables the construction of an interpolant that captures losing states at
%%%node $n$.

@@

We have discovered a set $\mathcal{I}$ of states losing for the environment.
Environment-losing states are only losing for a particular bound: given that
there does not exist an environment strategy that forces the game into an error
state in $k$ rounds or less; there may still exist a longer environment-winning
strategy.  We therefore record learned environment-losing states along with
associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.
$B^m[k]$ are initialised to $\top$ for all $k$.  Whenever an environment-losing
set $\mathcal{I}$ is discovered for a node $n$ with bound $\textsc{height}(k,
n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

@@

We also present a cactus plot of the number of instances solved over time (Figure~\ref{fig:cactus}). For
reference we plotted our results along with \textsc{Demiurge}~\cite{bloem2014},
the only SAT-based tool in the competition, the winner of the sequential
realisability track \textsc{Simple BDD Solver 2}~\cite{walker2014}, and
AbsSynthe (seq3) \cite{brenguier2014}. Our results differ from the competition,
we suspect due to the weaker hardware.
